---
title: My first experience with text mining
author: Antonio Schettino
date: '2018-01-29'
slug: text-analysis-wuthheights
categories:
  - R
tags:
  - R Markdown
  - text analysis
output:
  blogdown::html_page:
    toc: yes
    number_sections: no
summary: My first analysis of a document using ```tidytext```.
---

```{r setup, include = FALSE}

# ### install packages
# install.packages("knitr")
# install.packages("tidytext")
# install.packages("gutenbergr")
# install.packages("tidyverse")
# install.packages("ggthemes")
# install.packages("viridis")
# install.packages("reshape2")
# install.packages("wordcloud")
# install.packages("topicmodels")

### load packages
library(knitr)
library(tidytext)
library(gutenbergr)
library(tidyverse)
library(ggthemes)
library(viridis)
library(reshape2)
library(wordcloud)
library(topicmodels)

# for each chunk:
opts_chunk$set(
  echo = FALSE, # do not display the R code
  warning = FALSE, # no package warnings
  message = FALSE, # no package messages
  fig.width = 12, fig.height = 8 # figure size
)

# set seed for LDA model
seed.smorfia <- 67 # 'O purpo int'a chitarra
set.seed(seed.smorfia) # set seed for RNG

```

# Introduction

In this blog post I will engage in something a bit more serious than playing with [static](https://asch3tti.netlify.com/post/attractive-attractor/) or [dynamic](https://asch3tti.netlify.com/post/happy-holidays/) images (although, admittedly, it's a lot of fun). I will show you my first attempt at analyzing a document using the wonderful [```tidytext```](https://cran.r-project.org/web/packages/tidytext/index.html) package.   
   
```tidytext``` allows to sort the constituent elements of a text (e.g., single words or *n*-grams) following Hadley Wickham's [tidy principles](http://vita.had.co.nz/papers/tidy-data.pdf):

* each variable is a column;
* each observation is a row;
* each type of observational unit is a table.
   
It then becomes quite easy to extrapolate word frequencies (how often they occur) and word correlations (do specific words follow each other?), run sentiment analysis (overall positive/negative emotional connotation of the text), and topic modeling (i.e., identify words associated with a pre-specified number of overarching topics). All this can be done in a single document, across the body of work of an author, or large heterogeneous datasets (e.g., tweets).

# The book

For this exercise, I decided to explore **Wuthering Heights** by *Emily Brontë*. When I read it (a long, long time ago), I was not particularly hit by the troubled love between Heathcliff and Catherine -- although I was definitely attracted by the uncontainable, raw, and primordial vortex of passion that pervaded their relationship --, but rather by Heathcliff's quest for revenge against the upper class, especially embodied by Hindley and Edgar.   
   
Besides my appreciation for the novel, the choice of analyzing **Wuthering Heights** also came from practical considerations:

* it is available for free from the [Project Gutenberg](https://www.gutenberg.org/files/768/768-h/768-h.htm) collection, because copyright has expired;
* the full text can easily be downloaded with the [```gutenbergr```](https://cran.r-project.org/web/packages/gutenbergr/index.html) package;
* I could find several websites[^1] that refreshed my memory with respect to the main characters and events of the novel.
   
Following the instructions in Julia Silge and David Robinson's fantastic book [Text Mining with R](https://www.tidytextmining.com/tidytext.html), I was able to download the text, separate it into individual words, and eliminate stop words (e.g., "the", "of", "to") which are not informative. Then I decided to extract **word frequency**, perform **sentiment analysis**, and **topic modeling**.

# Word frequency

The first step was to quantify how often words were used across the 34 chapters of the novel, to have an initial idea of the content. So, I counted the number of occurrences for each word and selected only the most common ones (i.e., occurring at least 100 times).

```{r word_frequency}

WH.ID <- filter(gutenberg_metadata, author == "Brontë, Emily") # extract metadata for Emily Brontë's "Wuthering Heights"
WH <- gutenberg_download(WH.ID$gutenberg_id) %>% # download text
  unnest_tokens(word, text) %>% # split text into individual words
  anti_join(stop_words) # eliminate stop words (e.g., "the", "of", "to", ...)

# what are the most common words used in this book?
WH %>%
  count(word, sort = TRUE) %>% # count number of occurrences of each word
  filter(n >= 100) %>% # keep only words whose count is > 100
  mutate(word = reorder(word, n)) %>% # reorder words as a function of number of occurrences
  ggplot(aes(word, n)) + # plot
  geom_col(aes(fill = word), color = "black", size = .7, show.legend = FALSE) + # bars
  scale_y_continuous(breaks = seq(0, 450, 50)) + # x-axis: tick marks
  geom_hline(
    yintercept = seq(0, 450, 50), # reference lines
    linetype = "dotted", # line: type
    colour = "#999999", # line: color
    size = .8, # line: thickness
    alpha = .5 # line: transparency
  ) +
  geom_text(aes(label = n), # add numbers next to bars
    size = 5, # text size
    position = position_dodge(.9), vjust = .5, hjust = -.2 # text position
  ) +
  labs(
    title = "Word frequency in \"Wuthering Heights\"", # labels: title
    x = NULL, # labels: x-axis
    y = "number of occurrences" # labels: y-axis
  ) +
  coord_flip() + # flip axes
  scale_fill_viridis(option = "magma", discrete = TRUE) + # colors
  theme_pander(base_size = 18, pc = "white") # custom theme

```

Unsurprisingly, the name of the protagonist is the most frequent word in the novel, with `r as.double(count(WH, word, sort=TRUE)[1, "n"])` occurrences. The name of the co-star also pops up: *Catherine Earnshaw* (`r as.double(count(WH, word, sort=TRUE)[3, "n"])` occurrences). It's quite interesting to see the word *Linton* (`r as.double(count(WH, word, sort=TRUE)[2, "n"])` occurrences) between *Heathcliff* and *Catherine*: the two protagonists passionately love each other, but Catherine is seduced by the elegant manners of the Lintons and decides to marry *Edgar* (`r as.double(count(WH, word, sort=TRUE)[20, "n"])` occurrences).   
*Master* is another frequent word... why? Well, Heathcliff marries *Isabella Linton*, a calculated act of revenge that allows him to become the legal owner of Thrushcross Grange. Later, he forces *Cathy* (Catherine and Edgar's daughter; `r as.double(count(WH, word, sort=TRUE)[13, "n"])` occurrences) and *Linton* (Heathcliff and Isabella's son)[^2] to get married, thus becoming Cathy's father-in-law. When Edgar dies, Heathcliff becomes the master of Wuthering Heights.

# Sentiment analysis

This novel is dark. No wonder that it was received with mixed reviews when its first edition was published in 1847. The [Wikipedia](https://en.wikipedia.org/wiki/Wuthering_Heights) page on Wuthering Heights quotes a passage from a review originally published on *Graham's Lady Magazine*:

> "How a human being could have attempted such a book as the present without committing suicide before he had finished a dozen chapters, is a mystery. It is a compound of vulgar depravity and unnatural horrors."

This spicy quote obviously urged me to explore what kind of emotions are conveyed by this devilish work of moral corruption. I extracted sentiment classifications from the [NRC Word-Emotion Association Lexicon](http://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm) and selected four emotions: **anger**, **fear**, **joy**, and **sadness**.   
Here are the 10 most frequent words in each emotion class:

```{r sentiment_analysis}

# get sentiment classifications from the NRC lexicon
nrc.emo <- get_sentiments("nrc") %>%
  filter(sentiment == "joy" |
    sentiment == "anger" |
    sentiment == "fear" |
    sentiment == "sadness")

WH.emo <- WH %>%
  inner_join(nrc.emo) %>% # match words and sentiments
  count(word, sentiment) %>% # count occurrences while keeping sentiment information
  arrange(sentiment) # sort according to sentiment

# what are the most common words split into joy, anger, fear, and sadness?
WH.emo %>%
  group_by(sentiment) %>%
  top_n(10) %>% # take the top 10 words for each sentiment class
  ungroup() %>%
  mutate(word = reorder(word, n)) %>% # reorder words as a function of number of occurrences
  ggplot(aes(word, n, fill = sentiment)) + # plot
  geom_col(color = "black", size = .7, show.legend = FALSE) + # bars
  scale_y_continuous(breaks = seq(0, 100, 25)) + # x-axis: tick marks
  geom_hline(
    yintercept = seq(0, 100, 25), # reference lines
    linetype = "dotted", # line: type
    colour = "#999999", # line: color
    size = .8, # line: thickness
    alpha = .5 # line: transparency
  ) +
  geom_text(aes(label = n), # add numbers next to bars
    size = 5, # text size
    position = position_dodge(.9), vjust = .5, hjust = -.2 # text position
  ) +
  labs(
    title = "Top 10 words clustered by sentiment", # labels: title
    x = NULL, # labels: x-axis
    y = "number of occurrences" # labels: y-axis
  ) +
  scale_fill_viridis(option = "viridis", discrete = TRUE) + # colors
  facet_wrap(~sentiment, scales = "free_y") + # subplots according to sentiment
  coord_flip() + # flip axes
  theme_pander(base_size = 18, pc = "white") # custom theme

```

Indeed, tragedy seems to be omnipresent in the life of these characters. *Ill* occurs `r as.double(WH.emo[249, "n"])` times, *death* `r as.double(WH.emo[109, "n"])` times: Catherine Earnshaw "becomes ill" when emotionally upset (which happens fairly often) and dies right after giving birth to Cathy; Linton is a sickly boy; Edgar's physical states rapidly deteriorates, until the Grim Reaper finally claims his soul[^3]. The terror of physical ailments is palpable throughout the novel, and quite reasonably given the numerous [diseases](https://en.wikipedia.org/wiki/Diseases_and_epidemics_of_the_19th_century) spreading throughout Europe in the 19th century (cholera, typhus, smallpox, ...).   
We can also see words considered emotionally negative but actually used in neutral ways in the novel: for instance, *fire* is categorized as conveying fear in the NRC lexicon, but Emily Brontë mostly uses it to indicate a cooking fire, a cosy fireplace, or candle light. A funny misattribution is also evident with the word *mistress*, obviously used as a form of courtesy in Wuthering Heights but rated as conveying anger in the NRC lexicon because... because... because BDSM? No idea.   
Finally, the word *love* occurs `r as.double(WH.emo[1251, "n"])` times. The NRC lexicon classifies it as a joyful word but I wouldn't be so sure in this case.   
   
Just for fun, I also made a word cloud with the top 50 words categorized as positive or negative according to the lexicon by [Bing Liu and collaborators](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html).

```{r wordcloud, fig.height = 6, fig.width = 4}

WH %>%
  # match words and sentiments, this time from the sentiment lexicon of Liu et al. (positive/negative classes)
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  group_by(sentiment) %>%
  top_n(50) %>% # take the top 50 words for each sentiment class
  ungroup() %>%
  mutate(word = reorder(word, n)) %>% # reorder words as a function of number of occurrences
  acast(word ~ sentiment, value.var = "n", fill = 0) %>% # comparison.cloud requires matrix
  comparison.cloud(colors = c("#ab3400", "#469280")) # cloud representation of occurrence of positive and negative words

```

# Topic modeling

Finally, I used [Latent Dirichlet Allocation (LDA)](http://videolectures.net/mlss09uk_blei_tm/), a popular unsupervised Bayesian classification method, to divide the terms into 2 sections, hoping to somehow pick up the different events occurring during Heathcliff's childhood as opposed to his later years.

```{r LDA}

# prepare data for topic modeling
WH.DocumentTermMatrix <- WH %>%
  count(word, sort = TRUE) %>% # count number of occurrences of each word
  mutate(
    book = "Wuthering Heights", # add book name (necessary for cast_dtm)
    word = reorder(word, n) # reorder words as a function of number of occurrences
  ) %>%
  cast_dtm(book, word, n) # one term per row

# Latent Dirichlet Allocation
WH.LDA <- LDA(WH.DocumentTermMatrix,
  k = 2, # before/after Heathcliff runs away from Wuthering Heights?
  control = list(seed = seed.smorfia) # set seed for RNG
)

WH.LDA.topics <- tidy(WH.LDA, matrix = "beta") %>% # extract beta coefficients from model
  group_by(topic) %>%
  top_n(10, beta) %>% # take the top 10 words for each topic
  ungroup() %>%
  mutate(
    term = reorder(term, beta), # reorder terms according to beta values
    topic = recode(factor(topic), # rename topics
      "1" = "Heathcliff's childhood",
      "2" = "Heathcliff's maturity"
    )
  )

ggplot(WH.LDA.topics, aes(term, beta, fill = factor(topic))) + # plot
  geom_col(color = "black", size = .7, show.legend = FALSE) + # bars
  scale_y_continuous(breaks = seq(0, .015, .005)) + # x-axis: tick marks
  geom_hline(
    yintercept = seq(0, .015, .005), # reference lines
    linetype = "dotted", # line: type
    colour = "#999999", # line: color
    size = .8, # line: thickness
    alpha = .5 # line: transparency
  ) +
  labs(
    title = "Top 10 words clustered by topic", # labels: title
    x = NULL, # labels: x-axis
    y = expression(beta) # labels: y-axis
  ) +
  scale_fill_viridis(option = "viridis", discrete = TRUE) + # colors
  facet_wrap(~topic, scales = "free_y") + # subplots according to topic
  coord_flip() + # flip axes
  theme_pander(base_size = 18, pc = "white") # custom theme

```

*Heathcliff*, *Catherine*, and the *Lintons* (especially *Edgar*) are the major characters in this book, so it's obvious to find their names in both topics. The *Earnshaw* family seems to be more frequent in the first topic, which makes sense. In particular, the role of Mr. Earnshaw is pivotal: he is the one who travels to Liverpool, finds and adopts the orphan Heathcliff, and brings him to Wuthering Heights.   
Besides characters' names, some words seem to be more probable in one topic compared to the other. For example, *master* is more common in topic 2, because Heathcliff becomes the master of both Wuthering Heights and Thrushcross Grange later on in the novel.   
So, the LDA algorithm might have picked up the intended subdivision, although it may work much more efficiently when subdividing several documents into topics (because of the larger amount of data).   
   
Alright, this was a long post but I had a lot of fun exploring this classical English novel. There are so many other amazing things that you can do with ```tidytext```... if you're curious, check out [Text Mining with R](https://www.tidytextmining.com/tidytext.html)!

***
***

Below you will find the code to reproduce my first attempt at text mining... try with your favorite book, or expand this process by including all the works of your preferred writer!

```{r text_analysis_wuthheights, eval = FALSE, echo = TRUE}

# load relevant packages
library(tidytext)
library(gutenbergr)
library(tidyverse)
library(reshape2)
library(wordcloud)
library(topicmodels)
library(ggthemes)
library(viridis)

WH.ID <- filter(gutenberg_metadata, author == "Brontë, Emily") # extract metadata for Emily Brontë's "Wuthering Heights"
WH <- gutenberg_download(WH.ID$gutenberg_id) %>% # download text
  unnest_tokens(word, text) %>% # split text into individual words
  anti_join(stop_words) # eliminate stop words (e.g., "the", "of", "to", ...)

##################################################################################
############################ WORD FREQUENCY ######################################
##################################################################################

# what are the most common words used in this book?
WH %>%
  count(word, sort = TRUE) %>% # count number of occurrences of each word
  filter(n >= 100) %>% # keep only words whose count is > 100
  mutate(word = reorder(word, n)) %>% # reorder words as a function of number of occurrences
  ggplot(aes(word, n)) + # plot
  geom_col(aes(fill = word), color = "black", size = .7, show.legend = FALSE) + # bars
  scale_y_continuous(breaks = seq(0, 450, 50)) + # x-axis: tick marks
  geom_hline(
    yintercept = seq(0, 450, 50), # reference lines
    linetype = "dotted", # line: type
    colour = "#999999", # line: color
    size = .8, # line: thickness
    alpha = .5 # line: transparency
  ) +
  geom_text(aes(label = n), # add numbers next to bars
    size = 5, # text size
    position = position_dodge(.9), vjust = .5, hjust = -.2 # text position
  ) +
  labs(
    title = "Word frequency in \"Wuthering Heights\"", # labels: title
    x = NULL, # labels: x-axis
    y = "number of occurrences" # labels: y-axis
  ) +
  coord_flip() + # flip axes
  scale_fill_viridis(option = "magma", discrete = TRUE) + # colors
  theme_pander(base_size = 18, pc = "white") # custom theme

##################################################################################
############################ SENTIMENT ANALYSIS ##################################
##################################################################################

# get sentiment classifications from the NRC lexicon
nrc.emo <- get_sentiments("nrc") %>%
  filter(sentiment == "joy" |
    sentiment == "anger" |
    sentiment == "fear" |
    sentiment == "sadness")

WH.emo <- WH %>%
  inner_join(nrc.emo) %>% # match words and sentiments
  count(word, sentiment) %>% # count occurrences while keeping sentiment information
  arrange(sentiment) # sort according to sentiment

# what are the most common words split into joy, anger, fear, and sadness?
WH.emo %>%
  group_by(sentiment) %>%
  top_n(10) %>% # take the top 10 words for each sentiment class
  ungroup() %>%
  mutate(word = reorder(word, n)) %>% # reorder words as a function of number of occurrences
  ggplot(aes(word, n, fill = sentiment)) + # plot
  geom_col(color = "black", size = .7, show.legend = FALSE) + # bars
  scale_y_continuous(breaks = seq(0, 125, 25)) + # x-axis: tick marks
  geom_hline(
    yintercept = seq(0, 125, 25), # reference lines
    linetype = "dotted", # line: type
    colour = "#999999", # line: color
    size = .8, # line: thickness
    alpha = .5 # line: transparency
  ) +
  geom_text(aes(label = n), # add numbers next to bars
    size = 5, # text size
    position = position_dodge(.9), vjust = .5, hjust = -.2 # text position
  ) +
  labs(
    title = "Top 10 words clustered by sentiment", # labels: title
    x = NULL, # labels: x-axis
    y = "number of occurrences" # labels: y-axis
  ) +
  scale_fill_viridis(option = "viridis", discrete = TRUE) + # colors
  facet_wrap(~sentiment, scales = "free_y") + # subplots according to sentiment
  coord_flip() + # flip axes
  theme_pander(base_size = 18, pc = "white") # custom theme

# wordcloud
WH %>%
  # match words and sentiments, this time from the sentiment lexicon of Liu et al. (positive/negative classes)
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  group_by(sentiment) %>%
  top_n(50) %>% # take the top 50 words for each sentiment class
  ungroup() %>%
  mutate(word = reorder(word, n)) %>% # reorder words as a function of number of occurrences
  acast(word ~ sentiment, value.var = "n", fill = 0) %>% # comparison.cloud requires matrix
  comparison.cloud(colors = c("#ab3400", "#469280")) # cloud representation of occurrence of positive and negative words

##################################################################################
############################ TOPIC MODELING ######################################
##################################################################################

# prepare data for topic modeling
WH.DocumentTermMatrix <- WH %>%
  count(word, sort = TRUE) %>% # count number of occurrences of each word
  mutate(
    book = "Wuthering Heights", # add book name (necessary for cast_dtm)
    word = reorder(word, n) # reorder words as a function of number of occurrences
  ) %>%
  cast_dtm(book, word, n) # one term per row

# Latent Dirichlet Allocation
WH.LDA <- LDA(WH.DocumentTermMatrix,
  k = 2, # before/after Heathcliff runs away from Wuthering Heights?
  control = list(seed = seed.smorfia) # set seed for RNG
)

WH.LDA.topics <- tidy(WH.LDA, matrix = "beta") %>% # extract beta coefficients from model
  group_by(topic) %>%
  top_n(10, beta) %>% # take the top 10 words for each topic
  ungroup() %>%
  mutate(
    term = reorder(term, beta), # reorder terms according to beta values
    topic = recode(factor(topic), # rename topics
      "1" = "Heathcliff's childhood",
      "2" = "Heathcliff's maturity"
    )
  )

ggplot(WH.LDA.topics, aes(term, beta, fill = factor(topic))) + # plot
  geom_col(color = "black", size = .7, show.legend = FALSE) + # bars
  scale_y_continuous(breaks = seq(0, .015, .005)) + # x-axis: tick marks
  geom_hline(
    yintercept = seq(0, .015, .005), # reference lines
    linetype = "dotted", # line: type
    colour = "#999999", # line: color
    size = .8, # line: thickness
    alpha = .5 # line: transparency
  ) +
  labs(
    title = "Top 10 words clustered by topic", # labels: title
    x = NULL, # labels: x-axis
    y = expression(beta) # labels: y-axis
  ) +
  scale_fill_viridis(option = "viridis", discrete = TRUE) + # colors
  facet_wrap(~topic, scales = "free_y") + # subplots according to topic
  coord_flip() + # flip axes
  theme_pander(base_size = 18, pc = "white") # custom theme

```

[^1]: Many thanks to [wuthering-heights.co.uk](wuthering-heights.co.uk), [SparkNotes](http://www.sparknotes.com/lit/wuthering/), and the omnipresent [Wikipedia](https://en.wikipedia.org/wiki/Wuthering_Heights).
[^2]: Indeed, the high frequency of "Linton" may not only refer to the Linton family, but also Heathcliff and Isabella's son... damn, Miss Brontë, why did you have to repeat the same names over and over again?!?
[^3]: There is also another character who is ill: Lockwood, a wealthy British man who rents Thrushcross Grange, meets Heathcliff at Wuthering Heights, freaks out when he dreams about Catherine Earnshaw's ghost, and then gets ill. During his convalescence, he is told the whole story by Nelly Dean, the housekeeper of Thrushcross Grange.


